Variable Selection in Convex Function Estimation

### 2013-10-02 tex/scam/appendix.tex
updated proof of additive faithfulness

Let $\ds f^*_1,...,f^*_s \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$

Then $f^*_k(x_k) = \E[ f(X) \given x_k]$.
\end{lemma}

@@

Let $f^*_1,...,f^*_s$ be the minimizers as defined. It must be then that $f^*_k$ minimizes $\{ \E | f(X) - \sum_{k' \neq k} f^*_{k'} (X_{k'}) - f_k (X_k)|^2 \,:\, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

Fix $x_k$, we will show that the value $\E[ f(X) | x_k]$, for all $x_k$,  uniquely minimizes 
\[
\min_{ f_k(x_k) } \int_{\mathbf{x}_{-k}} p(x) | f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) - f_k (x_k)|^2 d \mathbf{x}_{-k}.
\]
If this is true, then the function $x_k \mapsto \E[ f(X) \given x_k]$ is the unique $f^*_k(x_k)$ that minimizes the expected square error.

Take the derivative with respect to the value $f_k(x_k)$ and set it equal to zero, we get that
\begin{align*}
\int_{\mathbf{x}_{-k}} p(\mathbf{x}) f_k(x_k) d \mathbf{x}_{-k} &= \int_{\mathbf{x}_{-k}} p(\mathbf{x}) ( f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) ) d \mathbf{x}_{-k} \\  
   p(x_k) f_k(x_k) &= p(x_k) \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) ( f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) ) d \mathbf{x}_{-k} \\
f_k(x_k) &= \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f(\mathbf{x}) d \mathbf{x}_{-k} 
 \end{align*}
Where the second line follows because $p(\mathbf{x}) = \prod_k p(x_k)$ by product distribution assumption. The third line follows because $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f^*_{k'}(x_{k'}) d\mathbf{x}_{-k} = 0$ for all $k' \neq k$ by the constraint that $\E[f^*_{k'}(X_{k'})] = 0$.

The square error objective is strongly convex. The second derivative with respect to $f_k(x_k)$ is $2 p(x_k)$, which is always positive under the assumption that $p$ is positive. Therefore, the solution $f^*_k(x_k) = \E[ f(X) \given x_k ]$ is unique.

Now, we verify that as a function of $x_k$, $\E[ f(X) | x_k]$ has mean zero and is convex. The former is true because $\E f(X) = 0$; the latter is true because for every $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k})$ is a convex function with respect to $x_k$ and therefore, $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}$ is still convex.

@@

Let $f^*_1,...,f^*_s \coloneqq \arg\min \{ \E | f(X) - \sum_k f_k(X_k) |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$.

The following are equivalent:
\begin{packed_enum}
\item $f$ does not depends on coordinate $k$
\item For all $x_k$, $\E[ f(X) | x_k] = 0$.
\end{packed_enum}
\end{proposition}

@@

The first condition trivially implies the second because $\E f(X) = 0$.

Fix $k$. Suppose that, for all $x_k$, $\E[ f(X) | x_k] = 0$.

By the assumption that , we know that, for all $x_k$,
\begin{align*}
p(x_k) \E[ f(X) | x_k ] &=  \int_{\mathbf{x}_{-k}} p(\mathbf{x_{-k}}) f(x_k,\mathbf{x}_{-k}) d \mathbf{x}_{-k} =0
\end{align*}

@@

For all $x_k, \mathbf{x}_{-k}$, 
\[
f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k \geq 0
\]
and
\[
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) (f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k) d\mathbf{x}_{-k} = 0
\]
Since $f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k$ is a continuous function of $\mathbf{x}_{-k}$ and the density $p(\mathbf{x}_{-k})$ is, by assumption, non-zero on the support, we conclude that for all $x_k, f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k =0 $ necessarily and thus $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k$.

@@

\begin{lemma}

\end{lemma}

By Theorem~\ref{thm:convex_faithful}, $R_s(f^{*s}) - R_s(f^{*(s-1)}) \geq \alpha > 0$.

@@

\begin{theorem} 
Suppose assumptions A1', A2', A3 hold. Suppose $f_0$ depends on all $s$-variables.

>>>>

Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq \arg\min \{ \E | f(X) - \sum_{k=1}^s f_k(X_k) - \mu |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$

Then $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and $\mu^* = \E f(X)$ and this solution is unique.
\end{lemma}

@@

Let $f^*_1,...,f^*_s, \mu^*$ be the minimizers as defined.

We first show that the optimal $\mu^* = \E f(X)$ for any $f_1, ..., f_k$ such that $\E f_k(X_k) = 0$. This follows from the stationarity condition, which states that $\mu^* = \E[ f(X) - \sum_k f_k(X_k)] = \E[ f(X) ]$.\\

We now turn our attention toward the $f^*_k$'s.

It must be then that $f^*_k$ minimizes $\{ \E | f(X) - \mu^* - \sum_{k' \neq k} f^*_{k'} (X_{k'}) - f_k (X_k)|^2 \,:\, f_k \trm{ convex }, \E f_k(X_k) = 0 \}$.

Fix $x_k$, we will show that the value $\E[ f(X) | x_k] - \mu^*$, for all $x_k$,  uniquely minimizes 
\[
\min_{ f_k(x_k) } \int_{\mathbf{x}_{-k}} p(x) | f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'}) - f_k (x_k)|^2 d \mathbf{x}_{-k}.
\]
It easily follows then that the function $x_k \mapsto \E[ f(X) \given x_k] - \mu^*$ is the unique $f^*_k$ that minimizes the expected square error.

Take the derivative with respect to the value $f_k(x_k)$ and set it equal to zero, we get that
\begin{align*}
\int_{\mathbf{x}_{-k}} p(\mathbf{x}) f_k(x_k) d \mathbf{x}_{-k} &= 
  \int_{\mathbf{x}_{-k}} p(\mathbf{x}) 
      ( f(\mathbf{x})-\sum_{k' \neq k} f^*_{k'}(x_{k'})-\mu^*) d \mathbf{x}_{-k} \\  
p(x_k) f_k(x_k) &= p(x_k) \int_{\mathbf{x}_{-k}} 
       p(\mathbf{x}_{-k}) ( f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})-\mu^*) 
              d \mathbf{x}_{-k} \\
f_k(x_k) &= \int_{\mathbf{x}_{-k}} 
       p(\mathbf{x}_{-k}) (f(\mathbf{x}) -\mu^*) d \mathbf{x}_{-k} 
 \end{align*}
Where the second line follows because $p(\mathbf{x}) = \prod_k p(x_k)$ by product distribution assumption. The third line follows because $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f^*_{k'}(x_{k'}) d\mathbf{x}_{-k} = 0$ for all $k' \neq k$ by the constraint that $\E[f^*_{k'}(X_{k'})] = 0$.

The square error objective is strongly convex. The second derivative with respect to $f_k(x_k)$ is $2 p(x_k)$, which is always positive under the assumption that $p$ is positive. Therefore, the solution $f^*_k(x_k) = \E[ f(X) \given x_k ] - \E f(X)$ is unique.

Now, we verify that as a function of $x_k$, $\E[ f(X) | x_k] - \E f(X)$ has mean zero and is convex. The former is clearly true; the latter is true because for every $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k})$ is a convex function with respect to $x_k$ and therefore, $\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) f(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}$ is still convex.

@@

Let $f^*_1,...,f^*_s \coloneqq \arg\min \{ \E | f(X) -\E f(X) - \sum_k f_k(X_k) |^2 \,:\, \forall k, f_k \trm{ convex }, \E f_k(X_k) = 0\}$.

The following are equivalent:
\begin{packed_enum}
\item $f$ does not depends on coordinate $k$
\item For all $x_k$, $\E[ f(X) | x_k] = \E f(X) $.
\end{packed_enum}
\end{proposition}

@@

The first condition trivially implies the second.

Fix $k$. Suppose that, for all $x_k$, $\E[ f(X) | x_k] = \E f(X)$.

By the assumption that , we know that, for all $x_k$,
\begin{align*}
 \E[ f(X) | x_k ] &=  \int_{\mathbf{x}_{-k}} p(\mathbf{x_{-k}}) f(x_k,\mathbf{x}_{-k}) d \mathbf{x}_{-k} = \E f(X)
\end{align*}

@@

For all $x_k, \mathbf{x}_{-k}$, 
\[
f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k \geq 0
\]
and
\[
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k}) (f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k) d\mathbf{x}_{-k} = 0
\]
Since $f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k$ is a continuous function of $\mathbf{x}_{-k}$ and the density $p(\mathbf{x}_{-k})$ is, by assumption, non-zero on the support, we conclude that for all $x_k, f(x_k, \mathbf{x}_{-k}) - f(0, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) x_k =0 $ necessarily and thus for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) = f(0, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) x_k$.

@@

\begin{lemma}
Suppose assumptions A1' and A4 hold.

Then $R(f^{*(s-1)}) - R(f^{*s}) \geq \alpha$, where $\alpha$ lower bounds the norm of the population optimal additive components as defined in assumpation A4.
\end{lemma}

\begin{align*}
R(f^{*(s-1)}) - R(f^{*s}) &= \E(f^{*(s-1)}(X) - f_0(X))^2 - \E(f^{*s}(X) - f_0(X))^2 \\
    &= \E( f^{*(s-1)}(X) - f^{*s}(X) + f^{*s}(X) - f_0(X))^2 - \E(f^{*s}(X) - f_0(X))^2 \\
    &= \E( f^{*(s-1)}(X) - f^{*s}(X))^2 - 2 \E[ (f^{*(s-1)}(X) - f^{*s}(X)) (f^{*s}(X) - f_0(X) ) ]
\end{align*}

@@

\begin{theorem} 
Suppose assumptions A1', A2', A3, A4 hold.

### 2013-10-02 tex/scam/icml14/optimization_icml.tex
updated proof of additive faithfulness

We now consider the following nonparametric regression problem
\begin{equation}\nonumber
          Y_{i} = f(\bds{x}_{i}) + \epsilon_{i} = 
                  \sum_{k=1}^{p}f_{k}(x_{ki}) + \epsilon_{i} \quad i=1,2,\cdots,n
\end{equation}
where $\bds{x}_{i}\in\mathbb{R}^{p}$ is the covariate, $Y_{i}$ is the
response and $\epsilon_{i}$ is mean zero noise. The regression function $f(\cdot)$ is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
a univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      h_{kj} \geq h_{ki} + \beta_{ki}(x_{kj}-x_{ki}) \quad (\forall i,j)
\end{equation}
where $h_{ki}\coloneqq f_{k}(x_{ki})$ and $\beta_{ki}$ is the
subgradient at point $x_{ki}$. We apparently need $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints, which is computationally
expensive for large scale problems.  In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to our optimization
program:
\begin{equation}
\begin{split}
       \min_{\bds{h},\bds{\beta},\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( Y_{i}-\sum_{k=1}^{p}h_{ki} - \mu \Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty} \\
       \textrm{subject to} &\;\; h_{k(i+1)} = h_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}), \\
                     & \;\; \sum_{i=1}^{n}h_{ki}=0,\\
                     & \;\; \beta_{k(i+1)} \geq \beta_{k(i)} \ (\forall k, i)
\end{split}
\label{np}
\end{equation}
Here $\{(1),(2),\ldots,(n)\}$ is a reordering of $\{1,2,\ldots,n\}$ such that $x_{k(1)}\leq{}x_{k(2)}\leq\cdots\leq{}x_{k(n)}$.  We introduce a mean parameter $\mu \in \R$ because $f$ may not have
zero mean. We can in fact solve for $\mu$ explicitly, as  
$\mu = \frac{1}{n} \sum_{i=1}^n Y_i = \bar{Y}$ which follows from the
KKT conditions
and the constraints $\sum_i h_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the supporting hyperplane constraints, as
\begin{align*}
  \forall j\geq{}i,\; h_{k(j)}-h_{k(i)}  &= 
            \sum\limits_{t=i}^{j-1}(h_{k(t+1)}-h_{k(t)}) \\
      & = \sum\limits_{t=i}^{j-1}\beta_{k(t)}(x_{k(t+1)}-x_{k(t)})\\ 
      & \geq \beta_{k(i)}\sum\limits_{t=i}^{j-1}(x_{k(t+1)}-x_{k(t)}) \\
      & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}) \\
  \forall j<i,\;  h_{k(j)}-h_{k(i)} &=
                \sum\limits_{t=j}^{i-1}(h_{k(t)}-h_{k(t+1)}) \\
     &= \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) \\ 
     & \geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)})  \\
     &= \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}
The $\ell_\infty/\ell_1$ penalty
$\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty}$ encourages group
sparsity of the vectors $\bds{\beta}_{k\cdot}$, and thus performs
variable selection.  We refer to this framework as the sparse convex
additive model (SCAM). While one can use
supporting hyperplanes to the epigraph as in \eqref{eq:outer}, 
SCAM uses the \emph{inner  piece-wise linear function}
that approximates the graph with secant lines. Notice that if we replace $\beta_{k(i+1)} \geq
\beta_{k(i)}$ with $\beta_{k(i+1)}=\beta_{k(i)}$, the optimization
reduces to the lasso.

>>>>

We now consider the following nonparametric regression problem
\begin{equation}\nonumber
          Y_{i} = f(\bds{x}_{i}) + \epsilon_{i} = 
                  \sum_{k=1}^{p}f_{k}(x_{ki}) + \epsilon_{i} \quad i=1,2,\cdots,n
\end{equation}
where $\bds{x}_{i}\in\mathbb{R}^{p}$ is the covariate, $Y_{i}$ is the
response and $\epsilon_{i}$ is mean zero noise. The regression function $f(\cdot)$ is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
a univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      h_{kj} \geq h_{ki} + \beta_{ki}(x_{kj}-x_{ki}) \quad (\forall i,j)
\end{equation}
where $h_{ki}\coloneqq f_{k}(x_{ki})$ and $\beta_{ki}$ is the
subgradient at point $x_{ki}$. We apparently need $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints, which is computationally
expensive for large scale problems.  In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to our optimization
program:
\begin{equation}
\begin{split}
       \min_{\bds{h},\bds{\beta},\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( Y_{i}-\sum_{k=1}^{p}h_{ki} - \mu \Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty} \\
       \textrm{subject to} &\;\; h_{k(i+1)} = h_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}), \\
                     & \;\; \sum_{i=1}^{n}h_{ki}=0,\\
                     & \;\; \beta_{k(i+1)} \geq \beta_{k(i)} \ (\forall k, i)
\end{split}
\label{np}
\end{equation}
Here $\{(1),(2),\ldots,(n)\}$ is a reordering of $\{1,2,\ldots,n\}$ such that $x_{k(1)}\leq{}x_{k(2)}\leq\cdots\leq{}x_{k(n)}$. We can solve for $\mu$ explicitly, as  
$\mu = \frac{1}{n} \sum_{i=1}^n Y_i = \bar{Y}$ which follows from the
KKT conditions
and the constraints $\sum_i h_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the supporting hyperplane constraints, as
\begin{align*}
  \forall j\geq{}i,\; h_{k(j)}-h_{k(i)}  &= 
            \sum\limits_{t=i}^{j-1}(h_{k(t+1)}-h_{k(t)}) \\
      & = \sum\limits_{t=i}^{j-1}\beta_{k(t)}(x_{k(t+1)}-x_{k(t)})\\ 
      & \geq \beta_{k(i)}\sum\limits_{t=i}^{j-1}(x_{k(t+1)}-x_{k(t)}) \\
      & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}) \\
  \forall j<i,\;  h_{k(j)}-h_{k(i)} &=
                \sum\limits_{t=j}^{i-1}(h_{k(t)}-h_{k(t+1)}) \\
     &= \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) \\ 
     & \geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)})  \\
     &= \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}
The $\ell_\infty/\ell_1$ penalty
$\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty}$ encourages group
sparsity of the vectors $\bds{\beta}_{k\cdot}$, and thus performs
variable selection.  We refer to this framework as the sparse convex
additive model (SCAM). While one can use
supporting hyperplanes to the epigraph as in \eqref{eq:outer}, 
SCAM uses the \emph{inner  piece-wise linear function}
that approximates the graph with secant lines. Notice that if we replace $\beta_{k(i+1)} \geq
\beta_{k(i)}$ with $\beta_{k(i+1)}=\beta_{k(i)}$, the optimization
reduces to the lasso.
